From Reversible Programs to \\ Univalent Universes and Back

### 2021-09-11 Pi/paper/introduction.tex
Fix "one generator" for free symmetric rig
Fixes #128

Such sequences of type isomorphisms are pervasive in reversible boolean circuits, which are at the core of quantum
computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
$k$~\cite{aaronson_et_al:LIPIcs:2017:8173,1201583}. But from the perspective of programming languages, permutations and
reversible circuits can be more conveniently expressed as isomorphisms over algebraic datatypes as proposed in the
$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012,theseus} whose syntax is inspired by the
sound and complete axiomatisation of type isomorphisms~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the
bicartesian structure, that is, coproducts and products. Since $\PiLang$ programs correspond to type isomorphisms of
finite types, our observation is that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid on
one generator.

It is folklore that the groupoid of finite sets and permutations is the free symmetric rig groupoid on one
generator~\cite{laplaza72,kelly74,baez2000finite}. Our main result formally establishes this correspondence by giving an
equational theory for $\PiLang$ that exactly includes all the necessary equations to decide equivalence of $\PiLang$
programs. As conjectured by \citet{caretteComputingSemiringsWeak2016}, these equations correspond to the \emph{coherence
  conditions of symmetric rig groupoids}, answering the conjecture in the positive.

>>>>

Such sequences of type isomorphisms are pervasive in reversible boolean circuits, which are at the core of quantum
computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
$k$~\cite{aaronson_et_al:LIPIcs:2017:8173,1201583}. But from the perspective of programming languages, permutations and
reversible circuits can be more conveniently expressed as isomorphisms over algebraic datatypes as proposed in the
$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012,theseus} whose syntax is inspired by the
sound and complete axiomatisation of type isomorphisms~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the
bicartesian structure, that is, coproducts and products. Since $\PiLang$ programs correspond to type isomorphisms of
finite types, our observation is that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid (on
zero generators).

It is folklore that the groupoid of finite sets and permutations is the free symmetric rig groupoid on zero
generators~\cite{laplaza72,kelly74,baez2000finite}. Our main result formally establishes this correspondence by giving an
equational theory for $\PiLang$ that exactly includes all the necessary equations to decide equivalence of $\PiLang$
programs. As conjectured by \citet{caretteComputingSemiringsWeak2016}, these equations correspond to the \emph{coherence
  conditions of symmetric rig groupoids}, answering the conjecture in the positive.

### 2021-09-11 Pi/paper/reversible1.tex
Fix "one generator" for free symmetric rig
Fixes #128

In reversible boolean circuits, the number of input bits matches the number of output bits. Thus, a key insight for a
programming language of reversible circuits is to ensure that each primitive operation preserves the number of bits,
which is just a natural number. The algebraic structure of natural numbers as the free commutative semiring (or,
commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides
sequential, vertical, and horizontal circuit composition operators.

>>>>

In reversible boolean circuits, the number of input bits matches the number of output bits. Thus, a key insight for a
programming language of reversible circuits is to ensure that each primitive operation preserves the number of bits,
which is just a natural number. The algebraic structure of natural numbers as the free commutative semiring (or,
commutative rig), with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides sequential, vertical, and
horizontal circuit composition operators.
